Walling up Backdoors in Intrusion Detection Systems

### 2019-10-16 paper/paper.tex
Some final small corrections

Pruning for neural networks has been proposed
%in the last century
as a method to simplify large neural networks \cite{sietsma_neural_1988}.
Pruning as a defense for neural networks against poisoning has emerged recently \cite{gu_badnets:_2017}. They proposed pruning as a defense mechanism against backdoored \glspl{cnn} and show that by removing infrequently used neurons from the last convolutional layer, potential backdoors can be removed. The rationale behind this is that some neurons specialize processing the regular samples while others focus on the backdoor.
To our knowledge, these pruning defences have been applied to \glspl{cnn} but not to \glspl{mlp}, which are commonly used for \glspl{ids} \cite{meghdouri_analysis_2018}. Although various pruning techniques have been proposed for \glspl{dt} in the last decades \cite{esposito_comparative_1997} with the aim of simplifying trees that overfit on the training data, pruning has not yet been investigated for its suitability for defending against backdoors for \glspl{dt} and \glspl{rf}.

@@

While we aim at sanitizing possibly backdoored \glspl{ids}, \cite{erlacher_how_2018} take a different approach: They create \textit{GENESIDS}, a tool that enables extensive testing of an \gls{ids}. This approach can potentially also uncover backdoors if it finds a case, in which the \gls{ids} surprisingly misbehaves.

\section{Experimental Setup} \label{sec:ml_approaches}
We performed our experiments with an \gls{rf} and an \gls{mlp} model and intentionally added a backdoor to both. In particular, we used the following experimental setup:
\paragraph{Datasets}
%\todo{Max: I would shorten here.}
%Several datasets for the purpose of building and evaluating IDSs have been developed. However, as pointed out in \cite{gharib_evaluation_2016}, there are numerous requirements that have to be met for a dataset to provide realistic performance benchmarks in this context.

@@

\paragraph{\gls{ml} Techniques}
%\subsubsection{Deep Learning}
We used PyTorch~\cite{paszke_automatic_2017} to build an \gls{mlp}.
% neural network classifier.
For this we used 5 fully connected layers with 512 neurons each. We applied \gls{relu} activation after each layer and dropout with a probability of 0.2 and used binary cross entropy  as loss function for training. % for the classification task itself.

@@

\paragraph{The Backdoor}

@@

In our case, we used the \gls{ttl} value contained in the \gls{ip} header of Internet traffic
as channel for signaling the backdoor.
%to signal the backdoor.
The \gls{ttl} is used for mitigating problems due to routing loops in IP networks.
%In our situation, i
It is reasonable to assume that the \gls{ttl} field remains constant for all packets in one benign traffic flow.
This assumption was perfectly reflected by the examined datasets, as only 965 flows exhibited a non-zero standard deviation for \cic{} and only 7 for \unsw{}.
%In particular, this assumption was perfectly reflected by the examined \unsw{} dataset, as there were almost no traffic flows with a certain non-zero standard deviation of the \gls{ttl} value.

@@

\paragraph{Performance Results}
\begin{table}[t]
\caption{Detection performance results.} \label{tab:performance_results}
\begin{tabular}{l r r r r} \toprule
& \multicolumn{2}{r}{\unsw{}} & \multicolumn{2}{r}{\cic{}} \\
& RF & DL & RF & DL \\ \midrule
%Accuracy	& 0.990 $\pm$ 0.000	& 0.989 $\pm$ 0.000	& 0.997 $\pm$ 0.000	& 0.998 $\pm$ 0.000	\\
%Precision	& 0.854 $\pm$ 0.001	& 0.845 $\pm$ 0.004	& 0.997 $\pm$ 0.000	& 0.999 $\pm$ 0.000	\\
%Recall	& 0.850 $\pm$ 0.003	& 0.829 $\pm$ 0.009	& 0.993 $\pm$ 0.000	& 0.992 $\pm$ 0.000	\\
%F1 score	& 0.852 $\pm$ 0.002	& 0.837 $\pm$ 0.003	& 0.995 $\pm$ 0.000	& 0.995 $\pm$ 0.000	\\
%Youden's J	& 0.845 $\pm$ 0.003	& 0.823 $\pm$ 0.009	& 0.992 $\pm$ 0.000	& 0.991 $\pm$ 0.000	\\
%Backdoor acc.	& 1.000 $\pm$ 0.000	& 0.998 $\pm$ 0.002	& 1.000 $\pm$ 0.000	& 1.000 $\pm$ 0.000	\\
Accuracy	& 0.990 & 0.989 & 0.997 & 0.998\\
Precision	& 0.854 & 0.845 & 0.997 & 0.999\\
Recall	& 0.850 & 0.829 & 0.993 & 0.992\\
F1 score	& 0.852 & 0.837 & 0.995 & 0.995\\
Youden's J	& 0.845 & 0.823 & 0.992 & 0.991\\
Backdoor acc.	& 1.000 & 0.998 & 1.000 & 1.000\\
\bottomrule
\end{tabular}
\end{table}
Table~\ref{tab:performance_results} shows performance results for
%both
the \gls{mlp} and \gls{rf}, depicting both detection performance of normal samples
%as contained in the respective datasets
and the efficacy of the backdoor. The models are thus able to detect the backdoor with high confidence while retaining high attack detection performance.
Our results are consistent with previous work like, e.g., \cite{meghdouri_analysis_2018}.

@@

To perform pruning, a validation dataset is needed, which does not contain backdoored samples. We take a validation set that is $\frac{1}{4}$ of the training set. We use the validation set for pruning as described in the next sections and a test set that is also $\frac{1}{4}$ to verify whether the backdoor can be removed successfully and how much the accuracy on the original data suffers. Training, validation and test sets are pairwise disjoint.

@@

Our experiments revealed that pruning does not remove the backdoor at all, while decreasing the accuracy for normal data if too many neurons are pruned. To check whether other reasons are responsible for the \gls{mlp}'s failure to remove the backdoor, we also conducted the following experiments: \begin{enumerate*} \item We did not take the average activation but the average of the binary activations of each neuron, which means that we did not consider the quantity of the activation but only if the activation is larger than 0 and then averaged the activations for all data samples. \item We checked whether the dropout regularization might be the reason. \item We hypothesized that the issue might be that there are a lot fewer malicious samples in the dataset than benign ones. Thus we reversed the backdoor and made the backdoor so that it would falsely classify benign samples as malicious ones.
\end{enumerate*}

@@

\subsubsection{Fine-Tuning}
%\todo{Max: It said pruning before but should be tuning, right?}
%AH. I guess both should work
\begin{figure}[b]
%\includegraphics[width=\columnwidth]{figures/finetuning_2015.pdf}
\includegraphics[width=\columnwidth]{figures/finetuning_2017.pdf}
\caption{Fine-tuning of the \gls{mui} for \unsw{}: While fine-tuning finally lowers backdoor accuracy a little bit, it takes more training epochs than training the original model.}
\label{fig:finetuning}
\end{figure}
In addition to pruning, we tried to use fine-tuning to remove the backdoor from the \gls{mlp}. % classifier.
%The use of
Fine-tuning exclusively makes  sense if both the computational effort and the required training set size for fine-tuning are substantially lower than for the original training procedure.

@@

\autoref{fig:pruningExample} shows an example of variant (3) of pruning applied to a tree
\begin{figure}[h]
\includegraphics[width=\columnwidth]{figures/prune_CAIA_backdoor_15/prune_oh_d.pdf}
\caption{Pruning an \gls{rf} for \unsw{}. The smaller the dashes, the smaller the validation dataset. The validation set size ranges from 1\% to 100\% of its original size.}
\label{fig:performancePruningRF}
\end{figure}
%\todo{Max: Which pruning algorithm actually?}
and \autoref{fig:performancePruningRF} shows the resulting accuracy for pruning according to variant (3) for the \unsw{} dataset. With sufficient leaves being pruned, the accuracy of the backdoor approaches zero. The accuracy for the regular data does not decrease significantly. It might even increase by reducing overfitting. With 90\% of leaves pruned, accuracy is still well above 99\%. Even with a very small validation set of just 1\% of its original size, the pruning still works as expected.

@@

%We found it
It is surprising that the neural network pruning and fine-tuning methods were ineffective for removing the backdoor from our \gls{dl} model in all our experiments. Since \glspl{cnn} are conceptually very similar to \glspl{mlp} it is not obvious that methods working for the former do not work for the latter. The difference that a \gls{cnn} always only looks at a portion of the input \cite{wikipedia_convolutional_2019} and not all of it (unlike an \gls{mlp}) should not change the efficacy of the pruning approach. This leads us to the conclusion that probably the proposed methods are insufficient for \glspl{mlp} and more research is required to develop methods suitable for them.

@@

>>>>

Pruning for neural networks has been proposed
%in the last century
as a method to simplify large neural networks \cite{sietsma_neural_1988}.
Pruning as a defense for neural networks against poisoning has emerged recently \cite{gu_badnets:_2017}. In \cite{gu_badnets:_2017} the authors proposed pruning as a defense mechanism against backdoored \glspl{cnn} and show that by removing infrequently used neurons from the last convolutional layer, potential backdoors can be removed. The rationale behind this is that some neurons specialize processing the regular samples while others focus on the backdoor.
To our knowledge, these pruning defences have been applied to \glspl{cnn} but not to \glspl{mlp}, which are commonly used for \glspl{ids} \cite{meghdouri_analysis_2018}. Although various pruning techniques have been proposed for \glspl{dt} in the last decades \cite{esposito_comparative_1997} with the aim of simplifying trees that overfit on the training data, pruning has not yet been investigated for its suitability for defending against backdoors for \glspl{dt} and \glspl{rf}.

@@

While we in this paper aim at sanitizing possibly backdoored \glspl{ids}, \cite{erlacher_how_2018} take a different approach: They create \textit{GENESIDS}, a tool that enables extensive testing of an \gls{ids}. This approach can potentially also uncover backdoors if it finds a case, in which the \gls{ids} surprisingly misbehaves.

\section{Experimental Setup} \label{sec:ml_approaches}
We performed our experiments with an \gls{rf} and an \gls{mlp} model and intentionally added a backdoor to both. In particular, we used the following experimental setup:
\subsection{Datasets}
%\todo{Max: I would shorten here.}
%Several datasets for the purpose of building and evaluating IDSs have been developed. However, as pointed out in \cite{gharib_evaluation_2016}, there are numerous requirements that have to be met for a dataset to provide realistic performance benchmarks in this context.

@@

\subsection{\gls{ml} Techniques}
%\subsubsection{Deep Learning}
We used PyTorch~\cite{paszke_automatic_2017} to build an \gls{mlp}.
% neural network classifier.
For this we used 5 fully connected layers with 512 neurons each. We applied \gls{relu} activation after each layer and dropout with a probability of 0.2 and used binary cross entropy  as loss function for training. % for the classification task itself.

@@

\subsection{The Backdoor}

@@

In our case, we used the \gls{ttl} value contained in the \gls{ip} header of Internet traffic
as channel for signaling the backdoor.
%to signal the backdoor.
The \gls{ttl} is used for mitigating problems due to routing loops in IP networks.
%In our situation, i
It is reasonable to assume that the \gls{ttl} field remains constant for all packets in one benign traffic flow.
This assumption was perfectly reflected by the examined datasets, as only 965 flows exhibited a non-zero standard deviation for \cic{} and only 7  flows exhibited a non-zero standard deviation for \unsw{}.
%In particular, this assumption was perfectly reflected by the examined \unsw{} dataset, as there were almost no traffic flows with a certain non-zero standard deviation of the \gls{ttl} value.

@@

\subsection{Performance Results}
\begin{table}[t]
\caption{Detection performance results.} \label{tab:performance_results}
\begin{tabular}{l r r r r} \toprule
& \multicolumn{2}{r}{\unsw{}} & \multicolumn{2}{r}{\cic{}} \\
& RF & DL & RF & DL \\ \midrule
%Accuracy	& 0.990 $\pm$ 0.000	& 0.989 $\pm$ 0.000	& 0.997 $\pm$ 0.000	& 0.998 $\pm$ 0.000	\\
%Precision	& 0.854 $\pm$ 0.001	& 0.845 $\pm$ 0.004	& 0.997 $\pm$ 0.000	& 0.999 $\pm$ 0.000	\\
%Recall	& 0.850 $\pm$ 0.003	& 0.829 $\pm$ 0.009	& 0.993 $\pm$ 0.000	& 0.992 $\pm$ 0.000	\\
%F1 score	& 0.852 $\pm$ 0.002	& 0.837 $\pm$ 0.003	& 0.995 $\pm$ 0.000	& 0.995 $\pm$ 0.000	\\
%Youden's J	& 0.845 $\pm$ 0.003	& 0.823 $\pm$ 0.009	& 0.992 $\pm$ 0.000	& 0.991 $\pm$ 0.000	\\
%Backdoor acc.	& 1.000 $\pm$ 0.000	& 0.998 $\pm$ 0.002	& 1.000 $\pm$ 0.000	& 1.000 $\pm$ 0.000	\\
Accuracy	& 0.990 & 0.989 & 0.997 & 0.998\\
Precision	& 0.854 & 0.845 & 0.997 & 0.999\\
Recall	& 0.850 & 0.829 & 0.993 & 0.992\\
F1 score	& 0.852 & 0.837 & 0.995 & 0.995\\
Youden's J	& 0.845 & 0.823 & 0.992 & 0.991\\
Backdoor acc.	& 1.000 & 0.998 & 1.000 & 1.000\\
\bottomrule
\end{tabular}
\end{table}
Table~\ref{tab:performance_results} shows performance results for
%both
the \gls{mlp} and \gls{rf}, depicting both detection performance of normal samples
%as contained in the respective datasets
and the efficacy of the backdoor. The models are thus able to detect the backdoor with high confidence while retaining high attack detection performance.
Our results are consistent with previous work like, e.g., \cite{meghdouri_analysis_2018}.

@@

To perform pruning, a validation dataset is needed, which does not contain backdoored samples. We take a validation set that is $\frac{1}{4}$ of the training set. We use the validation set for pruning as described in the next sections and a test set that is also $\frac{1}{4}$ of the training set to verify whether the backdoor can be removed successfully and how much the accuracy on the original data suffers. Training, validation and test sets are pairwise disjoint.

@@

Our experiments revealed that pruning does not remove the backdoor at all, while decreasing the accuracy for normal data if too many neurons are pruned. To check whether other reasons are responsible for the technique's failure to remove the backdoor, we also conducted the following experiments: \begin{enumerate*} \item We did not take the average activation but the average of the binary activations of each neuron, which means that we did not consider the quantity of the activation but only if the activation is larger than 0 and then averaged the activations for all data samples. \item We checked whether the dropout regularization might be the reason. \item We hypothesized that the issue might be that there are a lot fewer malicious samples in the dataset than benign ones. Thus we reversed the backdoor and made the backdoor so that it would falsely classify benign samples as malicious ones.
\end{enumerate*}

@@

\subsubsection{Fine-Tuning}
%\todo{Max: It said pruning before but should be tuning, right?}
%AH. I guess both should work
\begin{figure}[b]
%\includegraphics[width=\columnwidth]{figures/finetuning_2015.pdf}
\includegraphics[width=\columnwidth]{figures/finetuning_2017.pdf}
\caption{Fine-tuning of the \gls{mlp} for \unsw{}: While fine-tuning eventually lowers backdoor accuracy a little bit, it takes more epochs than training of the original model.}
\label{fig:finetuning}
\end{figure}
In addition to pruning, we tried to use fine-tuning to remove the backdoor from the \gls{mlp}. % classifier.
%The use of
Fine-tuning exclusively makes  sense if both the computational effort and the required training set size for fine-tuning are substantially lower than for the original training procedure.

@@

\autoref{fig:pruningExample} shows an example of variant (3) of pruning applied to a tree
\begin{figure}[b]
\includegraphics[width=\columnwidth]{figures/prune_CAIA_backdoor_15/prune_oh_d.pdf}
\caption{Pruning an \gls{rf} for \unsw{}. The smaller the dashes, the smaller the validation dataset. The validation set size ranges from 1\% to 100\% of its original size.}
\label{fig:performancePruningRF}
\end{figure}
%\todo{Max: Which pruning algorithm actually?}
and \autoref{fig:performancePruningRF} shows the resulting accuracy for pruning according to variant (3) for the \unsw{} dataset. With sufficient leaves being pruned, the accuracy of the backdoor approaches zero. The accuracy for the regular data does not decrease significantly. It might even increase by reducing overfitting. With 90\% of leaves pruned, accuracy is still well above 99\%. Even with a very small validation set of just 1\% of its original size, the pruning still works as expected.

@@

@@

%We found it
It is surprising that the neural network pruning and fine-tuning methods were ineffective for removing the backdoor from our \gls{dl} model in all our experiments. Since \glspl{cnn} are conceptually very similar to \glspl{mlp} it is not obvious that methods working for the former do not work for the latter. The difference that a \gls{cnn} always only looks at a portion of the input \cite{wikipedia_convolutional_2019} and not at all of it (unlike an \gls{mlp}) should not change the efficacy of the pruning approach. This leads us to the conclusion that probably the proposed methods are insufficient for \glspl{mlp} and more research is required to develop methods suitable for them.
